---
title: "Unveiling the EBA's Trade Triumph: Unmasking the Drivers of Export Boom in Least Developed Countries"
author: "Ahnaf Alam"
format:
  pdf:
    toc: true
    number-sections: true
date: today
date-format: long
bibliography: cite.bib
thanks: "The code and data is available at https://github.com/AhnafAlam1/Europe-and-apparel-export and the replicated files are available at https://www.socialsciencereproduction.org/reproductions/13e97e34-76f1-4768-b521-726c205262b6/index "
classoption: abstract
abstract: | 
  The European Union's Everything But Arms (EBA) trade scheme has led to increased exports for Least Developed Countries (LDCs), particularly in the apparel sector. However, existing studies don't conclusively attribute this export boom solely to the EBA. This paper fills this gap by examining the impacts of trade liberalization, factor endowments, and global demand on exports shock. The study confirms that EBA implementation increased exports, but recognizes other factors also played a crucial role. The paper implies that eliminating tariffs alone may not be enough to stimulate exports in developing nations and emphasizes addressing various additional factors affecting exports.
---
```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Writing a paper titled "Investigating the Impact of EBA on Export Surges in Least Developed Countries"
# Author: Ahnaf Alam
# Date: February 11, 2024
# Contact: ahnaf.alam@mail.utoronto.ca

#### Workplace setup ####
library(opendatatoronto)
library(tidyverse)
library(janitor)
library(knitr)
library(ggplot2)
library(ggpubr)
library(kableExtra)

#### Reading the data from Github file ####

bgd <- read_csv("outputs/data/bgd1.csv")
ldc <- read_csv("outputs/data/ldc22.csv")
m_s <- read_csv("inputs/data/market_share.csv")
nep <- read_csv("outputs/data/nep1.csv")
cmbd <- read_csv("outputs/data/cmbd1.csv")
comtrade <- read_csv("outputs/data/comtrade.csv")
```
# Introduction

In 2001, the European Union launched the Everything But Arms (EBA) initiative, granting tariff-and-quota-free market access to 47 of the world's least developed countries (LDCs) to boost development through increased exports [@faber2009everything]. The EBA exempted LDC imports from tariffs, except for arms, enhancing exports from Asian LDCs to the EU [@gradeva2010role]. However, existing studies don't conclusively attribute the export boom solely to the EBA, leaving uncertainties about the contributions of global demand, trade liberalization, and factor endowments to the trade shift in LDCs. The paper aims to address this gap.

European nations heavily depend on apparel imports; for example, Germany imports 95 percent of its apparel, while France, Italy, and Spain import 85 percent, 65 percent, and 55 percent, respectively [@gereffi2010global]. The EBA has sparked concerns about LDCs vying for trade contracts, leading to reduced clothing prices benefiting EU countries. Some studies criticize the scheme as "more political than developmental" [@page2002new]. European reliance on apparel imports is evident as trade preferences with Myanmar remained unaffected despite human rights concerns [@pennisi2023everything]. This underscores the role of European apparel imports in driving high export volumes in LDCs.

This paper uses @RAND's analysis on preferential market access, focusing on a Bangladesh case study, to establish how LDC exports evolved since the EBA's introduction. The original paper highlights how using internationally sourced textiles from LDCs leads to increased firm-level revenue growth for LDCs.

Utilizing graphs and figures, this paper demonstrates how LDCs capitalized on the EBA scheme, resulting in an export boom. However, it acknowledges that factors like trade liberalization, factor endowments, and global demand also influenced the export boost. The investigation concludes that while the EBA contributed to increased LDC exports, it is not the primary driver; other crucial factors play a role. This finding underscores the importance of global demand and trade liberalization for LDCs to benefit from free trade.

The paper begins with a discussion on the data sets used, followed by a Results section showcasing export increases after EBA. The subsequent Discussion section explores factors contributing to the overall trade level rise for all countries, EU and LDCs alike. Finally, the Conclusion section wraps up the paper with a discussion on improvements on the next iteration of this paper.

```{r}
#| label: tbl-1
#| tbl-cap: "Table explaining the variables in raw Eurostat data"
#| warning: false
#| message: false
#| echo: false

table1 <- 
  tibble(
    "Variable" = c("Declarant_lab", "Partner_lab", "Product", "Product_lab", "Stat_regime_lab", 
                   "Eligibility_lab", "Import_lab_regime", "year", "value"),
    "Description" = c("Importing country", "Exporting country", "HS Code. HS system is a standardized way of
                      classifing traded products", "Description of what is being traded", "How the imported 
                      product was processed", "Eligibility preferences of the product", "What type of prefereces
                      granted to the imported product", "Year of the trade", "value of the trade (in Euros)")
  )

table1 |> 
  kable(booktabs = TRUE) |>
  kable_styling(font_size = 9, latex_options = "hold_position")

# The last bit of the code was taken from paper by Chloe Thierson, titled "Catching Up With Republicans: 
# Democratic Marriage Rates Are on the Rise"
```

```{r}
#| label: tbl-2
#| tbl-cap: "Table displaying clean market share data set provided by Systma (2022). Source: UN Comtrade"
#| warning: false
#| message: false
#| echo: false

m_s$Year <- as.character(m_s$Year)
m_s[13:23, ] |>
  select(-1) |>
  kable(
    col.names = c("Year", "Export Value (in USD)", "Total trade (in USD)", "Share of trade", "China or LDCs"),
    digits = 2,
    booktabs = TRUE,
    linesep = "",
    format.args = list(big.mark = ",") 
  ) |>
  column_spec(c(1, 2, 3, 4, 5), width = "6em") |> 
  kable_styling(font_size = 9, latex_options = "scale_down")
```
# Data
R [@R], a statistical programming language, is exclusively used in this paper. For the purposes of cleaning and preparing the data, the libraries tidyverse [@tidy], janitor [@janet], knitr [@knitr], kableExtra [@kable] and wesanderson [@wes] are used. For the graphs, ggplot2 [@ggplot] and ggpubr [@gg2] is used.

## Incorporating LDC dataset

The raw data for `ldc` comes from Eurostat [@EU-EXTRA]. A description table for the raw data table is provided (see @tbl-1). The raw data was cleaned by removing all columns, excluding HS Code, value and year, which were used to create a new column, `util`. The new column reports on utilization rate of the EBA, which signifies the fraction of apparel imported under the EBA, relative to total apparel import from LDCs. Ultimately, the final market share data set includes three columns: year, util, and variance.  

Eurostat does not itself collect data. Rather, it relies on individual countries to report trade data through their statistical agencies [@2024]. The statistical agencies collect data from port customs. When a ship carrying imported product arrives at an EU country, the customs authorities record all the pertinent information regarding that shipment, including filling out all the variables in @tbl-1. Once compiled, each EU member states reports these data to the Eurostat for publishing.

## Mixing in Bangladesh, Cambodia and Nepal data

The `ldc` data set lacks country-specific information. To remedy this, country-specific data sets—bgd1 for Bangladesh, cmbd1 for Cambodia, and nep1 for Nepal—are generated by merging `ldc` data with raw data based on the year and column variables. These data sets provide information on the utilization rate and year for each country. The utilization rate now represents the fraction of EU apparel imported from a nation relative to the total EU apparel imported from that country [@RAND]. Additionally, the data separates between LDC exports of knit and woven materials.

## Information of UN Comtrade

The Market share data set, derived from UN Comtrade's WITS database [@WITS] and provided by Tobias Systma for replication. It focuses on export differences between China and LDCs in apparel exports from 2002 to 2018. The cleaned data includes columns for years, the value of apparel export, total export value, and the share of export—representing the ratio of apparel exports to total exports for both LDCs and China. For visualization, see @fig-1. 

WITS database is also utilized to gather information on LDCs’ apparel exports to high-earning OECD countries. This dataset mirrors the methodology and information used by market share dataset from Comtrade database. The dataset is limited to 2018 due to the COVID-19 pandemic, which led to the shutdown of non-essential trade.

Similar to Eurostat, the UN Comtrade database doesn't collect data directly; individual countries provide trade data. Both importing and exporting countries collect data through customs departments, including variables like HS Code, valuation in USD, and the importing country. If a country fails to provide data or if there are missing values, the UN Comtrade database estimates missing values by extrapolating from two adjacent periods [@un]. Missing gaps are filled every two weeks in the database.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-2
#| fig-cap: Utilization rate for apparel export for LDCs

ldc|>
  ggplot(mapping = aes(x = year, y = util)) +
  geom_path() +
  theme_classic() +
  labs(x = "Year",
       y = "Utilization Rate")

```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-1
#| fig-cap: Difference in market share for apparel export between China and LDCs between 2001 and 2018

m_s |>
  ggplot(mapping = aes(x = Year, y = share, group=type, color = type)) +
  geom_line(size = 2) +
  theme_minimal() +
  labs(x = "Year",
       y = "Percentage of market share (in decimals)") +
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45)) +
  scale_color_brewer(palette = "Accent", labels = c("China", "LDCs")) +
  guides(color = guide_legend(title = NULL))
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-3
#| fig-cap: Difference in knit vs woven material export for Bangladesh using EBA

p1 <- bgd |>
  ggplot(aes(x = year, y = frac, group = as.factor(woven), color = as.factor(woven))) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Year",
       y = "Utilization Rate") + 
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45)) +
  scale_color_brewer(palette = "Set1", labels = c("Woven", "Knitted"), breaks = c("1", "0")) +
  guides(color = guide_legend(title = NULL))
print(p1)
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-5
#| fig-cap: Difference in knit vs woven material export for Nepal using EBA


p2 <- nep |>
  ggplot(aes(x = year, y = frac, group = as.factor(woven), color = as.factor(woven))) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Year",
       y = "Utilization Rate") + 
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45)) +
  scale_color_brewer(palette = "Set1", labels = c("Woven", "Knitted"), breaks = c("1", "0")) +
  guides(color = guide_legend(title = NULL))
print(p2)
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-4
#| fig-cap: Difference in knit vs woven material export for Cambodia using EBA


p3 <- cmbd |>
  ggplot(aes(x = year, y = frac, group = as.factor(woven), color = as.factor(woven))) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Year",
       y = "Utilization Rate") + 
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45)) +
  scale_color_brewer(palette = "Set1", labels = c("Woven", "Knitted"), breaks = c("1", "0")) +
  guides(color = guide_legend(title = NULL))
print(p3)
```

# Results

## EBA Impact on LDC Exports

@fig-2 replicates Figure 1, illustrating the EBA utilization rate (2001-2018). This rate signifies the proportion of apparel imports from LDCs entering Europe under the EBA. Despite a generally increasing trend, a 20 percent uptick occurred in 2010-2011 [@EU-EXTRA], due to EBA policy changes relaxing local fabric requirements.

## China's Apparel Export Dominance

@tbl-2 summarizes the m_s dataset from [@UNComtrade], emphasizing differences in apparel exports between China and all LDCs. The table underscores substantial disparities in apparel export values, with China making up majority share of the apparel export industry. @fig-1, a reproduction of Figure 3, highlights this point. China's market share surged by 33 percent, compared to LDCs, which rose from 3 percent in 2002 to around 8 percent in 2018.

## Export Disparities: Woven vs. Knitted Materials

@fig-3, replicating Figure 4 of the original paper, depicts Bangladeshi apparel export utilization rates (2001-2018). The rate represents the ratio of EU imports under the EBA to total EU imports from Bangladesh, irrespective of the EBA. While knit apparel maintains a near 100 percent utilization rate, woven apparel saw a shift from roughly 20 percent in 2001 to nearly 100 percent by end-2018, with a notable increase in 2010-2011.

@fig-4 and @fig-5 extends the graphical framework in @fig-3, focusing on Cambodia and Nepal, respectively. Similarities include close to 100 percent utilization rates for both woven and knit apparel imports from 2011 onwards. Contrasting trends are evident, such as Cambodia consistently maintaining near-100 percent utilization for woven material, unlike Bangladesh and Nepal. Nepal and Bangladesh experienced a shift in woven apparel around 2010-2011, more pronounced in Bangladesh, while Cambodia experienced a shift in knit apparel imports around 2010-11.

## Apparel export to high-earning OECD countries

@fig-6 depicts value of LDC export to high-income OECD countries. The graph separates the knit material, with ProductCode 61 and ProductCode 62, for woven material. It shows that apparel exports from LDCs have been steadily increasing since 2001, with a slowdown in 2014. This slowdown is due to economic downturn at export destinations, causing apparel consumption to decrease by 7.92 percent [@islam_2016].

# Discussion

@fig-2 suggests that LDCs effectively utilized the EBA scheme, resulting in a substantial increase in the utilization rate post-implementation. Notably, Asian LDCs such as Bangladesh, Cambodia, and Nepal witnessed a surge in apparel exports in both knit and woven sectors. However, the market share of LDC exports significantly trailed behind China, as depicted in @fig-1. Nonetheless, the data leaves unanswered the question of whether the EBA was the exclusive reason for the upswing in LDC export levels or if other factors played a role. This is especially noticeable when we consider @fig-6, which points out how LDC exports increased to all high-earning OECD countries and not just to destinations that fall under the EBA scheme. This rationale suggests there must be additional underlying reasons that caused trade to increase. 

## Exploring Trade Liberalization's Impact on Export Growth

The notion of 'trade liberalization' emerges as a potential explanation for the notable increase in export levels in LDCs. This concept revolves around dismantling barriers to free trade, particularly tariffs that previously hindered poorer nations from competing with industrialized counterparts. The creation of the General Agreement on Tariffs and Trade (GATT) in 1947 marked a significant step in promoting trade liberalization. Since then, developing countries have experienced export growth, with manufacturing comprising 80 percent of their exports [@IMF_2001]. Studies show that trade liberalization led to a two-percentage-point increase in developing countries' exports since the 1990s [@santospaulino2002]. Free trade facilitates access to affordable inputs, creating export opportunities [@dornbusch1992]. Moreover, nations engaged in open trade tend to boast more diversified export structures compared to protectionist counterparts [@osakwe2018trade].

The concept of trade liberalization strongly suggests that the surge in LDC exports cannot be solely attributed to the EBA. It is evident that LDCs were already experiencing export growth before the EBA's implementation, owing to open and free trading relations with other nations.

## Global Demand as a Key Player in Export Dynamics

An essential factor influencing export levels is the global demand for resources. Global demand, especially during financial crises, tends to fluctuate, leading to price volatility adversely affecting a country's exports [@meyn2009implications]. Studies indicate that during global economic challenges, there is a substantial decrease in exports, with estimates suggesting a 20 percent decline in export due to financial factors [@amitiweinstein2011]. This phenomenon extends to LDCs, as financial crises in the global North often triggers reduced demand for exports from LDCs [@willenbockel2009global]. All these factors underscore how export levels, not confined solely to the global North, are susceptible to influences beyond schemes like the EBA.

## Factor endownments and export

A nation's factor endowments significantly influence its trade performance. Factor endowments encompasses resources such as land, capital, and natural resources available to a country. Research, including @coughlinfabel1988, indicates that this holds true for Least Developed Countries (LDCs), where both physical and human capital positively affect exports. @corvers1997explaining expand on this using the Heckscher-Ohlin-Vanek (HOV) model, which posits that countries export the factors they have in abundance and import scarce ones. Their study, integrating country-specific human capital endowments into the HOV model, reveals a positive impact of high-skilled labor and technological knowledge on trade [@corvers1997explaining]. This underscores the pivotal role of factor endowments in driving export growth.

## Limitations

The current data lacks information on trade sectors beyond apparel in trade. The Everything But Arms (EBA) agreement eliminates tariffs on imports, excluding arms. However, the absence of data on crucial sectors like agriculture and manufactured goods prevents the paper from establishing whether, post EBA implementation, only apparel exports increased or if a similar trend exists across all sectors. The subsequent version of this paper must integrate data from other sectors for a more robust analysis. This will help determine if EBA uniformly impacted trade positively or if the influence is specific to the apparel sector.

Although the paper acknowledges the roles of trade liberalization, global demand, and factor endowments in export demand, it lacks corresponding data. Furthermore, the investigation lacks regression models to confirm or rule out these factors as core reasons for the overall increase in export trade. Future researchers should address these gaps in the next version of the paper.
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-6
#| fig-cap: LDC exports to high-earning OECD countries for knit products vs woven products 


hs_61 <- comtrade |>
  filter(ProductCode == 61,
         Year <= 2018) |>
  ggplot(mapping = aes(x = Year, y = trade_in_1000USD)) +
  geom_line() +
  labs(title = "ProductCode 61", x = "Year", y = "Trade in 1000 USD")

hs_62 <-comtrade |>
  filter(ProductCode == 62,
         Year <= 2018) |>
  ggplot(mapping = aes(x = Year, y = trade_in_1000USD)) +
  geom_line() +
  labs(title = "ProductCode 62", x = "Year", y = "Trade in 1000 USD")

com <- ggarrange(hs_61, hs_62, nrow = 2)

print(com)

```
# Conclusion

The study looks to examine the impact of the Everything But Arms (EBA) policy on trade, with a focus on exports in the least developed countries (LDCs). Utilizing data from @RAND, the research confirms that exports, particularly in the apparel sector, have increased since the policy's initiation in 2001. However, the paper highlights that LDCs' apparel exports have not only risen to European countries but also to all advanced economies. Therefore, it can be inferred that external factors such as trade liberalization, global demand, and factor endowments may also contribute to the growth in exports. Consequently, attributing the export surge in LDCs solely to the EBA policy may not be sufficient. However, the study's limitation lies in the absence of regression analysis and data on sectors other than apparel, and that prevents us from definitively concluding that global demand, factor endowments and trade liberlization are the exclusive drivers of the trade shock.



\newpage
# Bibliography
